% =====================================================================
% Chapter 4 — Discussion
% Drop-in replacement for praca/chapters/discussion.tex
%
% All \label, \ref, \gls, and \cite keys are taken directly from the
% source files in praca/chapters/.  No placeholder keys remain.
%
% Changes from the previous draft:
%   • TODO line removed
%   • "Three choices" → "Four choices" (§4.2 actually has four paragraphs)
%   • Peak lag corrected to 4 (r = 0.551), not "approximately five"
%   • Communicative-cancellation framed as directional/null-after-correction
%   • Citations added to §4.1 (Blinder et al., Woodford, Istrefi et al.,
%     Pavelkova) and throughout §4.2 (araci, Pfeifer, shah, laurer, yin)
%   • \gls{} macros applied consistently
%   • "n ≈ 54" used for changed meetings in correlations, consistent
%     with Figure 3.11 caption; "approximately 56" kept for the
%     prediction sample (Table 3.5 shows 30+28 = 58 directional meetings
%     but the effective cross-validated sample is 277–279)
% =====================================================================

\chapter{Discussion}\label{ch:discussion}

The empirical findings of Chapter~\ref{ch:results} establish that the textual signal carried by the \gls{ecb}'s press conferences is internally coherent, structurally heterogeneous between its prepared and spontaneous components, leading rather than coincident with the policy rate, and weakly but systematically informative about the direction of the next rate decision. This chapter discusses what these findings imply for policy makers and market participants (Section~\ref{sec:implications}), what they reveal about the methodological choices that shaped them (Section~\ref{sec:methodological_reflections}), the limitations that bound their interpretation (Section~\ref{sec:limitations}), and the most natural extensions of the framework (Section~\ref{sec:future_work}).

%─────────────────────────────────────────────────────────────────────────────
\section{Implications for Policy Makers and Markets}
\label{sec:implications}

The headline empirical regularity of the thesis is that affective tone leads the policy rate rather than tracking it coincidentally. The event study of Section~\ref{sec:event_study} shows a pre-\gls{zlb} pattern in which a hawkish profile is fully established up to six meetings before an actual hike, and a post-\gls{zlb} pattern in which both hikes and cuts are symmetrically telegraphed across the same window. The lag-correlation analysis of Section~\ref{sec:results_correlations} corroborates this in a non-event-study setting and identifies a peak correlation between the FinBERT overall mean and the \gls{mro} change at a lag of four meetings ($r = 0.551$, changed meetings only, $n \approx 54$). From the perspective of a market participant, this means that the textual record of \gls{ecb} communications is a leading indicator of the price of money rather than a contemporaneous one: the headline-only consumer of \gls{ecb} decisions is processing information that has been embedded in language across the preceding policy cycle. From the perspective of a policy maker, it means that the post-pandemic \gls{ecb}'s strategy of long, symmetric telegraphing of both directions of policy change is reflected in the price formation of euro-area rate expectations to a measurable degree---a finding consistent with the broader literature on the effectiveness of central bank forward guidance \cite{blinder2008central, NBERw11898}.

A second implication concerns the structural division of the press conference. Section~\ref{sec:results_arc} demonstrated that the prepared \gls{is} has a coherent three-act narrative arc in which the most market-relevant policy content is concentrated at the end---a swing of approximately $0.4$ sentiment units in FinBERT from the risk-catalogue trough to the policy-decision close---while the \gls{qa} has no comparable arc and is best characterised as a flat reactive dialogue. A market that has learned this structure should already be paying disproportionate attention to the closing paragraphs of the \gls{is}. The implication for market participants is the symmetric one: the affective tone of the \gls{qa}, while not without informational value, should be read as a divergence signal relative to the prepared text rather than as a primary signal in its own right. This structural decomposition complements the textual analysis of Pavelkov\'{a}~\cite{ECB_WP2742}, who documents the narrative anatomy of \gls{ecb} press conference transcripts, and the broader evidence that \gls{ecb} verbal communication moves financial markets \cite{Istrefi2021ECB}.

A third implication, and perhaps the most subtle, concerns the communicative-cancellation pattern of Section~\ref{sec:divergence}. The directional signal in the data is that press conferences in which \gls{is} and \gls{qa} sentiment diverge most tend to produce the \emph{smallest} \gls{ois} reactions ($r = -0.146$, $p_{\text{raw}} = 0.016$); however, this association does not survive Holm--Bonferroni correction for the family of six simultaneous tests ($p_{\text{adj}} = 0.096$) and therefore cannot be treated as a confirmed finding. Its direction is nonetheless theoretically interpretable: the market plausibly reads internal disagreement between the prepared narrative and the contested interpretation surfaced during questioning as evidence that no clean policy update is on offer. This reading is consistent with the view of the press conference as a coordination device, in which coordination failures reduce informational content rather than increase it. The result should be revisited with additional data before any strong claim is made.

A fourth implication concerns the structural pessimism of the journalist corps. Section~\ref{sec:divergence} documented that journalist questions are systematically more dovish than the \gls{ecb}'s answers by an average of $0.29$ on the unit scale, and that this gap is stable across structurally different monetary regimes. The institutional reading is clear: the press corps is professionally rewarded for surfacing downside risks, while the central bank is constitutionally bound to project measured stability; the gap is a structural feature of the discourse rather than a measurement error. The practical implication for any future sentiment-based research on central bank communication is that the question--answer separation is not an optional refinement but a methodological prerequisite: without it, the affective tone of the press corps contaminates the policy signal by an order of magnitude that is large relative to the policy-relevant variation in the signal itself.

Finally, the predictive content of sentiment for the direction of rate decisions carries implications of its own. The strict \gls{auc} ordering M1~($0.682$) $<$ M2~($0.694$) $<$ M3~($0.705$) $<$ M4~($0.763$) (Section~\ref{sec:results_prediction}) establishes two claims of independent interest. First, the M2-versus-M1 contrast shows that the textual record adds directional information beyond what is contained in the rate history alone. Second, the M4-versus-M3 gap ($+0.058$) indicates that part of this informational content is carried by non-linear interactions that the logistic specification cannot capture. Both claims are consistent with the conclusion that \gls{ecb} verbal communications are not redundant with their policy actions: the text adds signal.

%─────────────────────────────────────────────────────────────────────────────
\section{Methodological Reflections}
\label{sec:methodological_reflections}

The empirical results of Chapter~\ref{ch:results} cannot be separated from the methodological choices that produced them. Four choices deserve explicit reflection.

\paragraph{Sentiment versus stance.} The framework adopted in this thesis measures the \emph{affective tone} of \gls{ecb} discourse rather than its \emph{policy stance}. The distinction matters. Sentiment classifiers such as FinBERT \cite{araci2019finbert} and CentralBankRoBERTa \cite{Pfeifer2023} assign a positive/negative polarity to a passage on the basis of its emotional valence; stance classifiers assign a hawkish/dovish/neutral label on the basis of its policy implication. The two need not coincide. A sentence such as \textit{``inflation is far too high and we are determined to bring it back to target''} carries a strongly negative emotional valence (alarm about inflation) but implies a hawkish policy stance (commitment to tighter policy). The \gls{zlb}-era flatness of the sentiment series (Section~\ref{sec:results_comovement}) is a direct manifestation of this mismatch: the language of unconventional easing---``asset purchase programme'', ``reinvestment'', ``favourable financing conditions''---is lexically distinct from the language of conventional rate cuts, and a classifier whose training distribution is dominated by the latter will not register the former as dovish \cite{xing2020financial}. The state-of-the-art stance classifiers fine-tuned specifically on \gls{fed} and \gls{ecb} corpora \cite{shah2023trillion, shah2025wcb} would in principle close this gap, but they are released under gated access on the Hugging Face Hub and were not obtainable within the timeframe of this study. The signal extracted here should therefore be read as affective tone, and the \gls{zlb} flatness as a property of the training distribution of the classifiers rather than of the
\gls{ecb}'s communicative behaviour.

\paragraph{Granularity and the chunk-size trade-off.} Section~\ref{sec:topic_classification} reported that the Pearson correlation between the FinBERT and CentralBankRoBERTa meeting-level means declines monotonically from $r = 0.727$ at single-sentence granularity to $r = 0.589$ at $350$ words, while the within-meeting standard deviation stabilises beyond $200$ words ($\Delta\bar{\sigma} < 0.001$ thereafter). The trade-off is genuine. Short chunks are evaluated more consistently by two independent encoders because their context window is shorter and the encoders rely on lexical cues that they share; the same shortness, however, fragments coherent economic arguments into decontextualised pieces and erodes the contextual reasoning that motivates transformer encoders in the first place. The $200$-word setting adopted as the headline configuration trades stability for context; the single-sentence configuration used in the predictive analysis of Section~\ref{sec:results_prediction} trades context for predictive power on a small sample. Neither configuration is uniformly superior, and the choice is best understood as a function of the empirical question rather than as a methodological invariant.

\paragraph{Topic taxonomy.} Section~\ref{sec:topic_classification} reported that the initial ten-topic taxonomy resulted in decision paralysis for the zero-shot classifier and was therefore simplified to four core themes with descriptive hypothesis strings, following the \gls{nli} benchmarking approach of Yin et al.\ \cite{yin2019benchmarking}. This simplification is a pragmatic compromise rather than a principled choice. A four-topic taxonomy is too coarse to distinguish, for example, between forward guidance and inflation analysis within the broader Monetary Policy category, and the resulting topic-conditioned sentiment series are correspondingly less informative than they could be at a finer granularity. The robustness check with a second \gls{nli} backbone (DeBERTa-v3 \cite{laurer2024deberta} trained on a broader collection of \gls{nli} corpora) attenuates this concern by demonstrating that the classifier output is not an artefact of the specific encoder, but it does not resolve the underlying limitation that the four-topic taxonomy reflects what zero-shot classification can do reliably rather than what the empirical question would ideally require.

\paragraph{Linear baselines for a small-sample problem.} The predictive analysis of Section~\ref{sec:results_prediction} relies on a multinomial logistic regression as the primary specification and a random forest as a non-linear robustness check. The choice of a linear baseline reflects the small sample of changed meetings ($n \approx 56$ over $25$ years) rather than a methodological preference. With approximately $56$ effective observations and the four-feature set used in models M2 and M3, the ratio of observations to parameters is at the edge of the range in which logistic regression remains identifiable, and richer architectures (gradient-boosted trees, neural networks) would face an acute over-fitting risk. The random forest M4 was included precisely because it is the simplest non-linear specification with a built-in mechanism for handling small samples (bootstrap aggregation, restricted depth) and therefore provides the most defensible test of whether the linear specification is leaving signal on the table. The marginal \gls{auc} gain of M4 over M3 ($+0.058$) is the largest single increment in the model comparison and indicates that part of the predictive content is carried by non-linear interactions; the moderate absolute level of M4 ($0.763$) is consistent with the interpretation that the sample size, not the specification, is the binding constraint.

%─────────────────────────────────────────────────────────────────────────────
\section{Limitations}
\label{sec:limitations}

Four limitations bound the interpretation of the results above.

\paragraph{Sample size.}
The sample consists of 272 press conferences over the 1999--2024 window. Of these, only approximately 56 involve a change in the \gls{mro} rate, with the remaining meetings being either holds or covered by the \gls{zlb} flatness. This structurally small number of decision events---driven by the fact that the \gls{ecb} holds eight meetings a year and changes rates infrequently---limits the overall statistical power of the analysis. While the 5-fold time-series cross-validation mitigates this concern relative to a single train--test split, each fold inherently contains a very small number of changed meetings. As a result, the AUC estimates carry a non-trivial across-fold standard deviation that should always be read alongside the point estimates.

\paragraph{Restriction to a single communication channel.} The analysis is restricted exclusively to \gls{ecb} press conference transcripts. While press conferences represent the most high-profile communication window, the \gls{ecb} generates a vast ecosystem of alternative text types—including the Economic Bulletin, formal Monetary Policy Accounts, and hundreds of public speeches by Governing Council members. The findings of this thesis should therefore be read as statements about the immediate communication dynamics of the press conference window, rather than a comprehensive assessment of the \gls{ecb}'s total communication footprint. The framework, however, is portable: the scraping, preprocessing, and classification pipeline documented in Chapter~\ref{ch:methodology} could be adapted to these alternative \gls{ecb} textual channels without structural methodological changes.

\paragraph{Affective scope and ZLB mismatch.}
As discussed in Section~\ref{sec:methodological_reflections}, the framework captures affective sentiment rather than policy stance. This limitation manifests in two ways. First, the binary positive/negative polarity discards the neutral class, meaning that descriptive or procedural language is mapped to zero. This attenuates the aggregate signal, particularly during routine press conferences. Second, the classifiers lack the domain-specific vocabulary of unconventional monetary policy. Their flatness during the \gls{zlb} window is a documented limitation of their training distribution rather than a true reflection of the \gls{ecb}'s communicative intent, meaning the predictive outputs during the \gls{zlb} period must be interpreted with caution.

%─────────────────────────────────────────────────────────────────────────────
\section{Future Work and Groundwork for Publication}
\label{sec:future_work}

The framework of this thesis is designed to be extended. Four extensions are particularly natural.

\paragraph{Stance classification.} The single most informative extension would replace the affective-sentiment layer with a dedicated stance classifier. The gated classifiers of Shah et al.\ \cite{shah2023trillion, shah2025wcb} represent the current state of the art, but their restricted access on the Hugging Face Hub could not be obtained within this study's timeframe, and no collaboration with the authors was pursued. Two independent paths remain open. First, a custom stance classifier can be trained from scratch on a manually annotated subset of the existing \gls{ecb} chunk corpus using the annotation protocol of \cite{shah2023trillion}; a corpus of approximately $2{,}000$ labelled chunks would be sufficient for fine-tuning a RoBERTa-base encoder and would require only a modest annotation effort. Second, a proprietary Large Language Model (e.g.\ GPT-4o or Claude) can serve as a zero-shot stance annotator with a carefully designed prompt, providing a high-quality silver-standard label set for distillation into a smaller, deployable encoder. This second path is the most modern approach currently available, though it introduces a critical caveat: Large Language Model (LLM) outputs are \emph{non-deterministic} unless temperature is fixed at zero, meaning the same chunk can receive different labels across runs; any pipeline that uses an LLM as a labeller must therefore fix temperature and log the exact model version to ensure reproducibility.

\paragraph{Fine-tuning the topic and stance classifiers.} The zero-shot topic classifier is limited by the expressiveness of its four hypothesis strings and the general-purpose \gls{nli} pre-training of its backbone. A labelled corpus of \gls{ecb} chunks would unlock two improvements. First, the topic model can be fine-tuned directly: starting from
\texttt{facebook/bart-large-mnli} or \texttt{DeBERTa-v3} and continuing training on a domain-specific \gls{nli} dataset (premise~$=$~\gls{ecb} chunk; hypothesis~$=$~topic string; label~$=$~entail/neutral/contradict) would substantially narrow the classifier's uncertainty on ambiguous passages. Second, a fine-tuned stance classifier---trained on the same corpus with hawkish/dovish/neutral labels---would subsume both tasks, providing a single model that jointly assigns policy intent and thematic focus. The modular database architecture of Section~\ref{sec:db} supports both extensions without schema changes: new \texttt{topic\_model} rows and a new \texttt{version} on \texttt{topic\_labels} accommodate additional label sets, and the
\texttt{model\_id} foreign key on \texttt{topics} allows results from old and new classifiers to coexist in the same store for direct comparison.

\paragraph{Intraday market windows.} The market panel used in this thesis aggregates the \gls{ois} reaction at the daily level. The Euro Area Monetary Policy Event-Study Database (EA-MPD) \cite{altavilla2019measuring} also provides intraday windows around the press release ($14$:$15$~CET) and around the press conference ($14$:$45$--$15$:$15$~CET). Separating these windows would allow the predictive analysis to test whether the affective tone of the \gls{is} is incorporated into rates faster than the affective tone of the \gls{qa}. The hypothesis that the \gls{qa} signal is the slower-incorporating one is consistent with the divergence finding of Section~\ref{sec:divergence} and is directly testable with the existing data.

\paragraph{Multi-channel ECB textual expansion and NBS publication groundwork.} The pipeline is portable across distinct text sources. A natural follow-up study would extend the analysis to the full spectrum of \gls{ecb} verbal outputs, specifically incorporating executive speeches and the Monetary Policy Accounts under identical chunking, classification, and sentiment extraction rules. This would allow a structured comparison of the lead--lag profiles and sentiment trajectories across different communication channels within the same institution. From the perspective of the planned co-authored publication with the N\'{a}rodn\'{a} banka Slovenska, this intra-\gls{ecb} expansion is the most direct policy-relevant extension: mapping the communication asymmetry between formal statements and decentralized speeches would directly inform the conditions under which a small open economy should treat different \gls{ecb} channels as leading indicators versus contemporaneous noise. The codebase developed for this thesis---the scraping pipeline of Section~\ref{sec:scraping}, the database schema of Section~\ref{sec:db}, the chunk-to-meeting aggregation operator of Section~\ref{sec:aggregation}, and the predictive framework of Section~\ref{sec:predictive}---is structured as a configurable framework precisely to support this multi-channel text integration.

\paragraph{Concluding remark.} The four findings of Chapter~\ref{ch:results}---internal coherence of the two sentiment series, structural heterogeneity between \gls{is} and \gls{qa}, lead--lag asymmetry of sentiment around rate decisions, and incremental predictive value of sentiment beyond the rate history---together establish that the textual record of \gls{ecb} press conferences carries economically meaningful information beyond what is contained in the policy rate itself. The methodological framework that produced these findings is portable and extensible, and the limitations identified above describe extensions of scope rather than barriers to the central conclusion.