\chapter*{Introduction} % chapter* je necislovana kapitola
\addcontentsline{toc}{chapter}{Introduction} % rucne pridanie do obsahu
\markboth{Introduction}{Introduction} % vyriesenie hlaviciek

The modern practice of monetary policy is, to a large extent, an exercise in communication. Since the abandonment of strategic opacity in the 1990s \cite{NBERw11898}, central banks have come to treat the management of expectations as a policy instrument in its own right \cite{blinder2008central, daly2025dynamic}. Each scheduled press conference of the \gls{ecb} is therefore a market-moving event: the textual record is not a description of past decisions but the channel through which the central bank shapes expected future rates, anchors inflation expectations, and influences financial conditions at horizons beyond its direct operational reach \cite{Istrefi2021ECB}.

The empirical study of these communications has progressed substantially along computational lines. Early work relied on dictionary-based methods that approximated tone by counting predefined positive and negative terms; the seminal observation of Loughran and McDonald \cite{LoughranMcDonald2011}---that words such as \textit{liability} carry no negative connotation in financial filings---made clear that general-purpose lexicons systematically misclassify domain-specific discourse. Transformer-based language models have since superseded this paradigm, evaluating each token in relation to its neighbours and resolving negation and qualification that bag-of-words methods cannot capture. In the central-bank domain, FinBERT \cite{araci2019finbert}---a BERT-base encoder fine-tuned on the Financial PhraseBank---and CentralBankRoBERTa \cite{Pfeifer2023}---a RoBERTa-base encoder fine-tuned on central-bank speeches---now define the operational standard for sentiment quantification, while recent stance classifiers \cite{shah2023trillion, shah2025wcb} and the BIS central-bank language models \cite{BIS_WP1215_2024} extend the frontier toward hawkish/dovish labels.

What the literature still routinely treats as a single object, however, is the press conference taken as a whole. Yet an \gls{ecb} press conference is in fact two structurally different communicative acts: a prepared \gls{is}, read by the President and exhaustively pre-agreed by the Governing Council, and an unscripted \gls{qa} session in which journalists set the agenda \cite{ECB_WP2742}. Treating the two as one document mixes the institution's official line with the structural pessimism of the press corps and obscures which phase carries the policy signal. The contribution of this thesis is to take that distinction seriously. We build a custom pipeline that isolates the \gls{is} from the \gls{qa} and separates journalists' questions from the President's answers, quantify the tone of each segment in parallel with FinBERT and CentralBankRoBERTa, and ask whether the resulting series carries information about the \gls{ecb}'s policy-rate cycle and the \gls{ois} curve---a market-implied term structure of expected future rates---over and above the rate history itself. Beyond its academic contribution, the work lays the empirical groundwork for a planned Národná banka Slovenska research project on the informational value of central-bank communications.

Three intuitions should be held in mind from the outset. First, the \gls{ecb} sets a short-term rate, but markets price an entire term structure of expected future rates, and the latter is what communication is designed to move. Second, sentiment (positive or negative tone) is not the same as stance (hawkish or dovish policy implication); the difference matters during episodes such as the \gls{zlb} period of unconventional easing, when the language of asset purchases is lexically distinct from that of conventional rate cuts. Third, the \gls{is} and the \gls{qa} carry structurally different information and must be analysed jointly but not interchangeably. Each idea receives formal treatment in the chapter that requires it.

The thesis is organised in four chapters. Chapter~\ref{ch:theory} establishes the theoretical background, covering the evolution of central-bank communication, the trajectory of \gls{nlp} techniques in financial text analysis, and the quantitative targets---policy rates, the Wu--Xia shadow rate, \gls{ois} rates, and structural shocks---against which the textual signal will be evaluated. Chapter~\ref{ch:data} documents the data and methodology: the scraping pipeline that produces the \gls{is}/\gls{qa}-split corpus of $289$ press conferences from June~1998 to December~2025, the chunk-level segmentation, the multi-model sentiment extraction, the zero-shot topic taxonomy, and the chunk-to-meeting aggregation. Chapter~\ref{ch:results} reports the empirical findings in four arcs of increasing demand on the data: classifier coherence and historical anchoring, the sectional anatomy of the press conference, sentiment around the policy-rate cycle, and the predictive value of the textual signal for the direction of the next rate decision. Chapter~\ref{ch:discussion} discusses the implications for policy makers and markets, reflects on the methodological choices, acknowledges the limitations, and identifies the most natural extensions---chief among them the cross-bank replication that would underpin the planned co-authored publication with the Národná banka Slovenska.