\chapter{Results}\label{ch:results}

This chapter reports the empirical findings of the thesis. The analysis proceeds in four arcs of increasing demand on the data. Section~\ref{sec:results_sentiment} establishes that the two sentiment classifiers produce internally coherent signals and shows that those signals track major historical episodes of monetary-policy stress. Section~\ref{sec:results_sectional} dissects the press conference into its prepared and spontaneous components, demonstrating that the \gls{is} and the \gls{qa} session carry structurally different information. Section~\ref{sec:results_cycle} relates the aggregate sentiment series to the \gls{ecb}'s policy-rate cycle, including an event-study of rate decisions. Section~\ref{sec:predictive} closes the loop by asking whether the textual signal carries predictive value for the direction of the next rate decision, comparing four classifiers of increasing complexity (a logit baseline, a sentiment-only logit, an augmented logit, and a random forest) and identifying the regimes in which sentiment-derived features add information beyond the rate history alone.

%─────────────────────────────────────────────────────────────────────────────
% actual sections:
\input{chapters/results/aggregate}
\input{chapters/results/sectional}
\input{chapters/results/sentiment_mro}
\input{chapters/results/prediction}

%─────────────────────────────────────────────────────────────────────────────

\paragraph{Summary of Chapter Findings.} Across the four arcs of analysis, four substantive findings have emerged. First, the two sentiment classifiers produce coherent, mutually validating signals that align with the major historical turning points of euro-area monetary policy. Second, the prepared \gls{is} and the spontaneous \gls{qa} are structurally different communicative acts, and the journalist--\gls{ecb} gap is a stable institutional feature that, if not separated by the pipeline, would have contaminated the policy signal by approximately $0.3$ on the unit scale. Third, sentiment leads the \gls{ecb}'s rate decisions by between two and ten meetings, with the lead being long and gradual in the pre-\gls{zlb} era and approximately symmetric in the post-\gls{zlb} hiking cycle. Fourth, sentiment features carry incremental predictive content beyond the rate history itself: AUC rises from $0.682$ (M1, rate-history only) to $0.694$ (M2, sentiment only) and $0.705$ (M3, augmented logit); a Random Forest on the same features adds a substantial further increment to $0.763$, indicating genuine non-linear interactions that the logistic specification cannot capture